Accelerating parameter inference with graphics processing units

### 2019-01-31 paper/paper.tex
paper.tex: spellcheck. References.

\section{Introduction}
The Advanced LIGO  \cite{2015CQGra..32g4001T}  and Virgo \cite{gw-detectors-Virgo-original-preferred}  ground-based gravitational wave (GW) detectors have
identified several coalescing compact binaries
\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}.  
%
The properties of the sources responsible have been inferred via Bayesian comparison of  each source with candidate
gravitational ave signals
\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}
\editremark{citations: real events, also LI and RIFT} \cite{gwastro-PE-AlternativeArchitectures,gwastro-PENR-RIFT}.  
Many more are expected as observatories reach design sensitivity \cite{2016LRR....19....1A}.  
Both to handle the rapid pace of discovery and particularly to enable coordinated multimessenger followup, GW
observatories should be able to reconstruct the full source parameters of coalescing binaries as fast as possible
\cite{LIGO-whitepaper-2015} \editremark{update}.    
However, particularly when using the best-available waveform models, these calculations can be very costly.  In
practice, detailed inferences even using simplified waveforms take weeks for binary neutron stars, and can take months
for costly time-domain effective-one-body models which incorporate the effects of precession.

% BORROWED FROM ROM: edit
Several strategies have been
developed to reduce the 
%overall computational cost and turnaround time of 
computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017arXiv170302062V}.  
% The computational cost of parameter estimation can be reduced 
Approaches that have appeared in the literature include
%by 
generating the approximate solutions more quickly \cite{gwastro-mergers-PE-ReducedOrder-2013,2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C,gwastro-mergers-IMRPhenomP,gwastro-SpinTaylorF2-2013}; 
%by 
interpolating some combination
of the waveform or likelihood \cite{gwastro-approx-ROMNR-Blackman2015,2013PhRvD..87d4008C,2013PhRvD..87l2002S,2014CQGra..31s5010P,2014PhRvD..90d4074S,gwastro-PE-AlternativeArchitectures,cole2014likelihood,2012MNRAS.421..169G};  or 
%by 
adopting a sparse representation to reduce the computational cost of data
handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures}.   % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
 Some methods, however, achieve rapid turnaround through simplifying approximations.  
%
The RIFT strategy described in \editremark{citations - include ROM} eschews these simplifications, performing
embarassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
significant net  computational cost and noticable startup time, limiting its diverse potential applications.

@@

Following \citet{gwastro-PE-AlternativeArchitectures}, we substitute expression~\eqref{eq:strain_mode} 
for $h_{\ell m}$ into the expression $h_k(t) =F_{+,k} h_+(t_k) +
  F_{\times,k}h_\times(t_k)$ for the detector response $h_k$, 
where $t_k=t_c - \vec{x}_k \cdot \hat{n}$ is the arrival time at the $k$th detector (at position $\vec{x}_k$)
for a plane wave propagating along $\hat{n}$ \cite{gwastro-PE-AlternativeArchitectures}.
We then substitute these expressions for $h_k$ into the likelihood function~\eqref{eq:loglikelihood}
thereby generating~\cite{gwastro-PE-AlternativeArchitectures}
\begin{widetext}
\begin{align}
\ln {\cal L}(\lambda, \theta) 
&= (D_{\rm ref}/D) \text{Re} \sum_k \sum_{\ell m}(F_k \Y{-2}_{\ell m})^* Q_{k,lm}(\lambda,t_k)\nonumber \\
&   -\frac{(D_{\rm ref}/D)^2}{4}\sum_k \sum_{\ell m \ell' m'}
\left[
{
|F_k|^2 [\Y{-2}_{\ell m}]^*\Y{-2}_{\ell'm'} U_{k,\ell m,\ell' m'}(\lambda)
}
% \right. \nonumber \\ & \left.
 {
+  \text{Re} \left( F_k^2 \Y{-2}_{\ell m} \Y{-2}_{\ell'm'} V_{k,\ell m,\ell'm'} \right)
}
\right]
\label{eq:def:lnL:Decomposed}
\end{align}
\end{widetext}
%% \noindent \scott{[FOR RICHARD: is eq 9/10 missing objects? Looks like $t_c$ an $F_k$ just appeared. On the other hand,
%%     putting them into earlier equations will complicate the discussion and is awkward.]}  
where 
 where $F_k = F_{+,k} - i F_{\times,k}$ are the
complex-valued detector
response functions of the $k$th detector \cite{gwastro-PE-AlternativeArchitectures} and 
the quantities $Q,U,V$ depend on $h$ and the data as
\begin{subequations}
%\label{eq:ComputeRhoViaInnerProductMatrix}
\label{eq:QUV}
\begin{align}
Q_{k,\ell m}(\lambda,t_k) &\equiv \qmstateproduct{h_{\ell m}(\lambda,t_k)}{d}_k \nonumber\\
&= 2 \int_{|f|>f_{\rm low}} \frac{df}{S_{n,k}(|f|)} e^{2\pi i f t_k} \tilde{h}_{\ell m}^*(\lambda;f) \tilde{d}(f)\ , \\
{ U_{k,\ell m,\ell' m'}(\lambda)}& = \qmstateproduct{h_{\ell m}}{h_{\ell'm'}}_k\ , \\
V_{k,\ell m,\ell' m'}(\lambda)& = \qmstateproduct{h_{\ell m}^*}{h_{\ell'm'}}_k  \ .
\end{align}
\end{subequations}
Rewriting sums as matrix operations, the likelihood can be equivalently expressed as
\begin{align}
\ln {\cal L} &= \frac{D_{\rm ref}}{D} \text{Re}[ (F Y)^\dag Q]  \nonumber \\
 & - \frac{D_{\rm ref}^2}{4 D^2} [ (FY)^\dag U FY + (FY)^TV FY]
\label{eq:lnL:MatrixForm}
\end{align}
where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ ar matricies over
$(l,m)$.

@@

We do benefit slightly by re-using the adaptive Monte Carlo intergrator for each extrinsic point $\lambda$.  Since the
integrator has already identified the likely range of sky locations and distances on the first iteration, each
subsequent evaluation of the marginalized likelihood can converge marginally more quickly.

@@

We illustrate the code using two synthetic signals: a binary black hole with masses $m_{1,2}=35,30 M_\odot$ and a binary
neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
Both signals are chosen to have peak SNR $\simeq 14$ and are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
spin.  We repeat this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
available in large numbers; (b) V100, available on selected high-performance machines; (c) and
GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.

\subsection{Massive compact binary black holes}
Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$
is associated with code setup followed by reading, data conditioning, and Fourier-transforming the
data.  The setup cost $\tau_{setup}$ arises from waveform generation and  the inner products $U,V,Q$.  Finally, the Monte Carlo
cost $\tau_{mc}=N_{ad} \tau_{ad}+N_{it}\tau_{eval}$ increases with the number of Monte Carlo iterations $N_{it}$ with
and the number of times $N_{ad}$ the sampling prior used in adaptive Monte Carlo is regenerated, in proportion to their
cost.   In the standard configuration, the adaptive sampling prior is regenerated every $n_{\rm chunk}$ iterations (i.e,
 $N_{ad} \simeq N_{it}/n_{chunk}$), regenerating a sampling prior in two sky location parameters and distance.  
The choice of one $\lambda$ for each instance of ILE was eue to the substantial time  $\tau_{MC}$, which vastly
dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version has
cost elements as shown in Table \ref{tab:CostBreakdown}, based on the assumption that the MC terminates after $N_{\rm it}\simeq 2\times 10^6$ iterations using
$n_{\rm chunk}\simeq 10^4$.  
The marginalized likelihood described above  converges incredibly rapidly to a small value if the candidate model is
inconsistent with a signal (or the absence thereof) in the data, with only $N_{it} \simeq O(10^4)$  evaluations needed.

@@

  hardware, for a two-mode system $(l,m)=\pm 2$ analysing $T=8\unit{s}$ of data with a massive binary black hole

@@

With the new low-cost ${\cal L}_{\rm margT}$ evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
comparable or in excess of the total time used to evaluate the marginalized likelihood ${\cal L}_{marg}$.   In fact, the total time per Monte Carlo
evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
carrying out scientific calculations by evaluationg the likelihood.  For these reasons, we reorganize the workflow, so each
instance of ILE loops over
$N_{eval}$ different choices for $\lambda$. 
%
Additionally, particularly for massive binary black holes, we investigate two additional performance improvements.  First and
foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
$\tau_{setup}$.  Previously and out of an abundance of caution, \ILE{} employed a sampling rate $1/\Delta t = 16384
\unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will the  cost of all operations with
timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
\emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
significantly-padded data buffer,
using 
``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
Reducing the duration of data analyzed by a factor $s'$ will reduce the cost $\tau_{setup}$ by a factor $s'$.  
%
Combining these factors together, the overall computational cost  $T_{ILE}/N_{eval}$  of each marginalized likelihood
evaluation is
\begin{align}
T_{ILE}/N_{\rm eval} &= \frac{\tau_{start}}{N_{\rm eval}} 
 + 
 [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
 ] 
\end{align}
For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
\editremark{numbers}.  
Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
increase with the overall number of samples, so the overall evaluation time scales as
\begin{align}
T_{ILE,mod}/N_{\rm eval} &= \frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
 \nonumber \\ &
 + 
 [ \tau_{setup}/s'  + N_{it}\tau_{it}
 ] 
\end{align}
Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
\unit{s}$ for a binary black hole.    Of this $25 \unit{s}$, only roughly
$1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
7.5\times 10^{-7}$.
Likewise, because overhead dominates over array manipulations, an analysis with  $16\unit{kHz}$ timeseries requires only
marginally more time than the corresponding $4\unit{kHz}$ analysis.

@@

The \ILE{} likelhood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1.2 \unit{s}$
at 4kHz and $1.9\unit{s}$ at 16 kHz.

The \ILE{} likelhood $\ln {\cal L}_{\rm margT}$ involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
the number of modes used.  This increased cost is most apparent in Table \ref{tab:CostBreakdown} when comparing the
columns corresponding  $\tau_{it,like}$ for GPU (b) between the $\pm 2$ and $(\pm 2,\pm 1)$ rows.  The cost of
evaluating $\ln {\cal L}_{\rm margT}$   with 4 modes is roughly twice that of runs with two modes, as expected.

@@

\subsection{Binary neutron stars}
\begin{table*}
\begin{tabular}{lrr|ccccc|rr}
Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ &
$\frac{T_{ILE}}{N_{eval}}$ & GPU \\  %\hline 
  &   Hz & m & sec & sec & & $\mu$sec & $\mu$sec  &sec  & use  \%\\ \hline 
% ./parse_report.sh 20190130-profile_bns_nogpu_16kHz_pcdev13.log | more
%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
%   (11261 - 10974)/20/100. = 0.14 sec for 16 kHz
CPU & 16384 & $\pm 2,\pm 1 $ & 35 & 26 & 0.14  & 680  & & 590    \\ 
       & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  & \\ \hline
%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
%   (226.489 - 17)/20/100  : 0.1 sec for 4kHz
%   (165 - 29.  )/20/100   : looks odd for 16 kHz
GPU (b) & 16384 & $\pm 2, \pm 1$   & 35 & 26 & 0.07 &  2.4 & &35 &  \\
       & 4096 & $ \pm 2, \pm 1 $        &  35  & 11.5 & 0.1 &  2.4  & &24 & \\ \hline
\end{tabular}
\caption{\label{tab:CostBreakdown:BNS}\textbf{Profiling performance: Binary neutron stars}: Evaluation costs for the
  marginalized likelihood on default
  hardware,  analysing $T=8\unit{s}$ of data with a binary neutron stars
  $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
}
\end{table*}

@@

\editremark{test on other hardware}

@@

%\bibstyle{unsrt}
\bibliography{paperexport}
\end{document}

>>>>

\section{Introduction}
The Advanced LIGO  \cite{2015CQGra..32g4001T}  and Virgo \cite{gw-detectors-Virgo-original-preferred}  ground-based gravitational wave (GW) detectors have
identified several coalescing compact binaries
\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}.  
%
The properties of the sources responsible have been inferred via Bayesian comparison of  each source with candidate
gravitational wave signals
\cite{DiscoveryPaper,LIGO-O1-BBH,2017PhRvL.118v1101A,LIGO-GW170814,LIGO-GW170608,LIGO-GW170817-bns}
\editremark{citations: real events, also LI and RIFT} \cite{gwastro-PE-AlternativeArchitectures,gwastro-PENR-RIFT}.  
Many more are expected as observatories reach design sensitivity \cite{2016LRR....19....1A}.  
Both to handle the rapid pace of discovery and particularly to enable coordinated multimessenger followup, GW
observatories should be able to reconstruct the full source parameters of coalescing binaries as fast as possible
\cite{LIGO-whitepaper-2015} \editremark{update}.    
However, particularly when using the best-available waveform models, these calculations can be very costly.  In
practice, detailed inferences even using simplified waveforms take weeks for binary neutron stars, and can take months
for costly time-domain effective-one-body models which incorporate the effects of precession.

% BORROWED FROM ROM: edit
Several strategies have been
developed to reduce the 
%overall computational cost and turnaround time of 
computational cost of parameter estimation \cite{gwastro-pe-Brandon-STF2,gwastro-PE-AlternativeArchitectures,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,2016PhRvD..94d4031S,2017CQGra..34k5006V}.  
% The computational cost of parameter estimation can be reduced 
Approaches that have appeared in the literature include
%by 
generating the approximate solutions more quickly \cite{gwastro-mergers-PE-ReducedOrder-2013,2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C,gwastro-mergers-IMRPhenomP,gwastro-SpinTaylorF2-2013}; 
%by 
interpolating some combination
of the waveform or likelihood \cite{gwastro-approx-ROMNR-Blackman2015,2013PhRvD..87d4008C,2013PhRvD..87l2002S,2014CQGra..31s5010P,2014PhRvD..90d4074S,gwastro-PE-AlternativeArchitectures,cole2014likelihood,2012MNRAS.421..169G};  or 
%by 
adopting a sparse representation to reduce the computational cost of data
handling \cite{antil2013two,gwastro-mergers-PE-ReducedOrder-2013,2016PhRvD..94d4031S,gw-astro-ReducedOrderQuadraturePE-TiglioEtAl2014,gwastro-PE-AlternativeArchitectures}.   % 2014PhRvX...4c1006F,2013PhRvD..87l2002S,2013PhRvD..87d4008C
 Some methods, however, achieve rapid turnaround through simplifying approximations.  
%
The RIFT strategy described in \editremark{citations - include ROM} eschews these simplifications, performing
embarrassingly-parallel inferences even for costly models.   However, the method as previously implemented still had a
significant net  computational cost and noticeable startup time, limiting its diverse potential applications.

@@

Following \citet{gwastro-PE-AlternativeArchitectures}, we substitute expression~\eqref{eq:strain_mode} 
for $h_{\ell m}$ into the expression $h_k(t) =F_{+,k} h_+(t_k) +
  F_{\times,k}h_\times(t_k)$ for the detector response $h_k$, 
where $t_k=t_c - \vec{x}_k \cdot \hat{n}$ is the arrival time at the $k$th detector (at position $\vec{x}_k$)
for a plane wave propagating along $\hat{n}$ \cite{gwastro-PE-AlternativeArchitectures}.
We then substitute these expressions for $h_k$ into the likelihood function~\eqref{eq:loglikelihood}
thereby generating~\cite{gwastro-PE-AlternativeArchitectures}
\begin{widetext}
\begin{align}
\ln {\cal L}(\lambda, \theta) 
&= (D_{\rm ref}/D) \text{Re} \sum_k \sum_{\ell m}(F_k \Y{-2}_{\ell m})^* Q_{k,lm}(\lambda,t_k)\nonumber \\
&   -\frac{(D_{\rm ref}/D)^2}{4}\sum_k \sum_{\ell m \ell' m'}
\left[
{
|F_k|^2 [\Y{-2}_{\ell m}]^*\Y{-2}_{\ell'm'} U_{k,\ell m,\ell' m'}(\lambda)
}
% \right. \nonumber \\ & \left.
 {
+  \text{Re} \left( F_k^2 \Y{-2}_{\ell m} \Y{-2}_{\ell'm'} V_{k,\ell m,\ell'm'} \right)
}
\right]
\label{eq:def:lnL:Decomposed}
\end{align}
\end{widetext}
%% \noindent \scott{[FOR RICHARD: is eq 9/10 missing objects? Looks like $t_c$ an $F_k$ just appeared. On the other hand,
%%     putting them into earlier equations will complicate the discussion and is awkward.]}  
where 
 where $F_k = F_{+,k} - i F_{\times,k}$ are the
complex-valued detector
response functions of the $k$th detector \cite{gwastro-PE-AlternativeArchitectures} and 
the quantities $Q,U,V$ depend on $h$ and the data as
\begin{subequations}
%\label{eq:ComputeRhoViaInnerProductMatrix}
\label{eq:QUV}
\begin{align}
Q_{k,\ell m}(\lambda,t_k) &\equiv \qmstateproduct{h_{\ell m}(\lambda,t_k)}{d}_k \nonumber\\
&= 2 \int_{|f|>f_{\rm low}} \frac{df}{S_{n,k}(|f|)} e^{2\pi i f t_k} \tilde{h}_{\ell m}^*(\lambda;f) \tilde{d}(f)\ , \\
{ U_{k,\ell m,\ell' m'}(\lambda)}& = \qmstateproduct{h_{\ell m}}{h_{\ell'm'}}_k\ , \\
V_{k,\ell m,\ell' m'}(\lambda)& = \qmstateproduct{h_{\ell m}^*}{h_{\ell'm'}}_k  \ .
\end{align}
\end{subequations}
Rewriting sums as matrix operations, the likelihood can be equivalently expressed as
\begin{align}
\ln {\cal L} &= \frac{D_{\rm ref}}{D} \text{Re}[ (F Y)^\dag Q]  \nonumber \\
 & - \frac{D_{\rm ref}^2}{4 D^2} [ (FY)^\dag U FY + (FY)^TV FY]
\label{eq:lnL:MatrixForm}
\end{align}
where $F,D$ are arrays over extrinsic parameters; $Q$ is an array over time and $(l,m)$; and $U,V$ are matrices over
$(l,m)$.

@@

We do benefit slightly by re-using the adaptive Monte Carlo integrator for each extrinsic point $\lambda$.  Since the
integrator has already identified the likely range of sky locations and distances on the first iteration, each
subsequent evaluation of the marginalized likelihood can converge marginally more quickly.

@@

We illustrate the code using two synthetic signals: a binary black hole with masses $m_{1,2}=35,30 M_\odot$ and a binary
neutron star with masses $1.4 M_\odot$ and $1.35 M_\odot$.
We generate  synthetic data for a two-detector LIGO configuration, assuming operation at  initial (BBH) or advanced (BNS) design  sensitivity.
Both signals are generated with SEOBNRv4.    We perform a multi-stage iterative RIFT analysis of these signal with the
time-domain SEOBNRv4 (BBH) or TaylorT4 (BNS) approximation, under the simplifying assumption that both objects are point particles with zero
spin.  We repeat this analysis on heterogeneous LVC collaboration computing resources, to assess our variable
performance across architectures.  Notably, we investigate the following GPU options: (a) GTX 1050 Ti at LIGO-CIT,
available in large numbers; (b) V100, available on selected high-performance machines; (c) and
GTX 1050 at LIGO-LHO.    Unless otherwise noted in the text, we will discuss code configurations using CPU-only and GPU (b) using
$4096\unit{Hz}$ sampling and only the  $\ell=2, m=\pm 2$ modes.  We also conservatively report computational cost for
parameters $\lambda$ which are close to or within the posterior and therefore, due to their consistency with the data,
have the best-determined extrinsic parameters and thus converge the most slowly.

\editremark{report on distance to each signal, and in particular $d_{\rm max}$ as that might impact performance too}
chosen to have peak SNR $\simeq 14$ and

\subsection{Massive compact binary black holes}
Previously, each instance of ILE examined one intrinsic point $\lambda$.  The overall cost of this ILE evaluation
involved two parts: a startup cost, a setup cost, and a Monte Carlo integration cost.  The  startup cost $\tau_{start}$
is associated with code setup followed by reading, data conditioning, and Fourier-transforming the
data.  The setup cost $\tau_{setup}$ arises from waveform generation and  the inner products $U,V,Q$.  Finally, the Monte Carlo
cost $\tau_{mc}=N_{ad} \tau_{ad}+N_{it}\tau_{eval}$ increases with the number of Monte Carlo iterations $N_{it}$ with
and the number of times $N_{ad}$ the sampling prior used in adaptive Monte Carlo is regenerated, in proportion to their
cost.   In the standard configuration, the adaptive sampling prior is regenerated every $n_{\rm chunk}$ iterations (i.e,
 $N_{ad} \simeq N_{it}/n_{chunk}$), regenerating a sampling prior in two sky location parameters and distance.  
The choice of one $\lambda$ for each instance of ILE was due to the substantial time  $\tau_{MC}$, which vastly
dominated the overall computational cost.   For example, for a typical analysis of a short signal --  a typical binary
black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$ --  this version has
cost elements as shown in Table \ref{tab:CostBreakdown}, based on the assumption that the MC terminates after $N_{\rm it}\simeq 2\times 10^6$ iterations using
$n_{\rm chunk}\simeq 10^4$.  
The marginalized likelihood described above  converges incredibly rapidly to a small value if the candidate model is
inconsistent with a signal (or the absence thereof) in the data, with only $N_{it} \simeq O(10^4)$  evaluations needed.

@@

  hardware, for a two-mode system $(l,m)=\pm 2$ analyzing $T=8\unit{s}$ of data with a massive binary black hole

@@

With the new low-cost ${\cal L}_{\rm margT}$ evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
comparable or in excess of the total time used to evaluate the marginalized likelihood ${\cal L}_{marg}$.   In fact, the total time per Monte Carlo
evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
carrying out scientific calculations by evaluating the likelihood.  For these reasons, we reorganize the workflow, so each
instance of ILE loops over
$N_{eval}$ different choices for $\lambda$. 
%
Additionally, particularly for massive binary black holes, we investigate two additional performance improvements.  First and
foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
$\tau_{setup}$.  Previously and out of an abundance of caution, \ILE{} employed a sampling rate $1/\Delta t = 16384
\unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will the  cost of all operations with
timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
\emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
significantly-padded data buffer,
using 
``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
Reducing the duration of data analyzed by a factor $s'$ will reduce the cost $\tau_{setup}$ by a factor $s'$.  
%
Combining these factors together, the overall computational cost  $T_{ILE}/N_{eval}$  of each marginalized likelihood
evaluation is
\begin{align}
T_{ILE}/N_{\rm eval} &= \frac{\tau_{start}}{N_{\rm eval}} 
 + 
 [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
 ] 
\end{align}
For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
\editremark{numbers}.  
Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
increase with the overall number of samples, so the overall evaluation time scales as
\begin{align}
T_{ILE,mod}/N_{\rm eval} &= \frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
 \nonumber \\ &
 + 
 [ \tau_{setup}/s'  + N_{it}\tau_{it}
 ] 
\end{align}
Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
\unit{s}$ for a binary black hole.    Of this $25 \unit{s}$, only roughly
$1.5\unit{s}$ corresponds to evaluating the likelihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
7.5\times 10^{-7}$.
Likewise, because overhead dominates over array manipulations, an analysis with  $16\unit{kHz}$ timeseries requires only
marginally more time than the corresponding $4\unit{kHz}$ analysis.

@@

The \ILE{} likelihood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1.2 \unit{s}$
at 4kHz and $1.9\unit{s}$ at 16 kHz.

The \ILE{} likelihood $\ln {\cal L}_{\rm margT}$ involves sums over modes, with the most expensive likelihood and setup operations (involving $Q$) growing linearly with
the number of modes used.  This increased cost is most apparent in Table \ref{tab:CostBreakdown} when comparing the
columns corresponding  $\tau_{it,like}$ for GPU (b) between the $\pm 2$ and $(\pm 2,\pm 1)$ rows.  The cost of
evaluating $\ln {\cal L}_{\rm margT}$   with 4 modes is roughly twice that of runs with two modes, as expected.

@@

\subsection{Binary neutron stars}
\begin{table*}
\begin{tabular}{lrr|ccccc|rr}
Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ &
$\frac{T_{ILE}}{N_{eval}}$ & GPU \\  %\hline 
  &   Hz & m & sec & sec & & $\mu$sec & $\mu$sec  &sec  & use  \%\\ \hline 
% ./parse_report.sh 20190130-profile_bns_nogpu_16kHz_pcdev13.log | more
%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
%   (11261 - 10974)/20/100. = 0.14 sec for 16 kHz
CPU & 16384 & $\pm 2,\pm 1 $ & 35 & 26 & 0.14  & 680  & & 590    \\ 
       & 4096 & $\pm 2, \pm 1$ & 35 & &  &   &&  & \\ \hline
GPU (a) & 16384 & $\pm 2, \pm 1$   & 35 &  &&&&   \\
       & 4096 & $ \pm 2, \pm 1 $        &  35  & &&&& 360 \\ \hline
%   - resampling time: take (mcsampler - likelihood), divide by 20=Neval, divide by number of adaptive steps (100)
%   (226.489 - 17)/20/100  : 0.1 sec for 4kHz
%   (165 - 29.  )/20/100   : looks odd for 16 kHz
GPU (b) & 16384 & $\pm 2, \pm 1$   & 35 & 26 & 0.07 &  2.4 & &35 &  \\
       & 4096 & $ \pm 2, \pm 1 $        &  35  & 11.5 & 0.1 &  2.4  & &24 & \\ \hline
\end{tabular}
\caption{\label{tab:CostBreakdown:BNS}\textbf{Profiling performance: Binary neutron stars}: Evaluation costs for the
  marginalized likelihood on default
  hardware,  analyzing $T=8\unit{s}$ of data with a binary neutron stars
  $m_1=1.4 M_\odot,M_2=1.35 M_\odot$, with convergence threshold $n_{\rm eff} > 50$.
}
\end{table*}

@@

\editremark{test on other hardware -- also problem, a lot of the speed improvement occurs because the initial grid is
  crap/too wide}

@@

%\bibstyle{unsrt}
\bibliography{paperexport,LIGO-publications}
\end{document}
